A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-08-29 manuscript/paper.Rmd

```{r setup, include=FALSE}
source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/setup.R"))

@@

```{r, results='asis' }
add_stable(
  read_rds(figure_file("implementations_table.rds", "03-method_characterisation"))[[params$table_format]],
"implementations",
paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::collapse("; "))
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "03-method_characterisation"),
  ref_id = "implementation_qc_overview", 
  caption_main = "Overview of the quality control results for every method.", 
  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_checks')}). Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score. Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation. This is represented in the plot as the height on the y-axis. Top: Average QC score for each method. Right: The average score of each quality control item. Shown into more detail are those items which had a score lower than 0.5.")
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("tree.svg", experiment_id = "09-user_guidelines"),
  ref_id = "user_guidelines", 
  caption_main = "Practical guidelines for method users.", 
  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
)
```

@@

## Real datasets
We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynbenchmark](http://www.github.com/dynverse/dynbenchmark)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample; gold standard) (`r ref("stable", "datasets")`). These datasets are deposited in Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].

```{r}
add_stable(
  read_rds(figure_file("datasets.rds","01-datasets/05-dataset_characterisation"))[[params$table_format]], 
  "datasets", 
  "Real datasets used in this study."
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "samplers.rds",
    "01-datasets/05-dataset_characterisation/3-synthetic")
  )[[params$table_format]], 
  "samplers", 
  "Distributions from which each parameter in the thermodynamic model was sampled."
)
```

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.svg", experiment_id="02-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
 "Edge flip score values for common trajectory topologies",
  ""
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "qc_checks.rds",
    "03-method_characterisation"
  ))[[params$table_format]], 
  "qc_checks", 
  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
)
```

@@

```{r "supplementary", echo=FALSE, results='asis'}
if (params$table_format == "html") 
  source(paste0(dynbenchmark::get_dynbenchmark_folder(), "/analysis/paper/sfigs_stables.R"))
write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
write_rds(figs, derived_file("figs.rds", "paper"))
```

>>>>

```{r setup, include=FALSE}
source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/setup.R"))

@@

```{r, results='asis' }
add_stable(
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation"))[[params$table_format]],
"implementations",
paste0("Overview of the trajectory inference methods included in this study. ", glue::glue_data(non_inclusion_reasons, "{footnote}: {long}") %>% glue::collapse("; "))
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("implementation_qc_overview.svg", experiment_id = "4-method_characterisation"),
  ref_id = "implementation_qc_overview", 
  caption_main = "Overview of the quality control results for every method.", 
  caption_text = glue::glue("Shown is the score given for each method on every item from our quality control score sheet ({ref('stable', 'qc_checks')}). Each aspect of the quality control was part of a category, and each category was weighted so that it contributed equally to the final quality score. Within each category, each aspect also received a weight depending on how often it was mentioned in a set of papers discussing good practices in tool development and evaluation. This is represented in the plot as the height on the y-axis. Top: Average QC score for each method. Right: The average score of each quality control item. Shown into more detail are those items which had a score lower than 0.5.")
)
```

@@

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = figure_file("tree.svg", experiment_id = "7-user_guidelines"),
  ref_id = "user_guidelines", 
  caption_main = "Practical guidelines for method users.", 
  caption_text = "As the performance of a method most heavily depends on the topology of the trajectory, the choice of TI method will be primarily influenced by the user's existing knowledge about the expected topology in the data. We therefore devised a set of practical guidelines, which combines the method's performance, user friendliness and the number of assumptions a user is willing to make about the topology of the trajectory. Methods to the right are ranked according to their performance on a particular (set of) trajectory type. Further to the right are shown the user friendliness scores (++: ≥ 0.95, +: ≥ 0.85, ± ≥ 0.75, - ≥ 0.65), overall performance (++: top method, +: difference between top method's performance ≥ -0.05, ±: ≥-0.2, -: ≥ -0.5) and required prior information."
)
```

@@

## Real datasets
We gathered `r sum(tasks_info$task_source == "real")` real datasets by searching for "single-cell" at the Gene Expression Omnibus and selecting those datasets in which the cells are sampled from different stages in a dynamic process (`r ref("stable", "datasets")`). The scripts to download and process these datasets are available on our repository ([github.com/dynverse/dynalysis](http://www.github.com/dynverse/dynalysis)). Whenever possible, we preferred to start from the raw counts data. These raw counts were all normalised and filtered using a common pipeline, discussed later. We determined a reference standard for every dataset using labelling provided by the author's, and classified the standards into gold and silver based on whether this labelling was determined by the expert using the expression (silver standard) or using other external information (such as FACS or the origin of the sample; gold standard) (`r ref("stable", "datasets")`). These datasets are deposited in Zenodo ([doi.org/10.5281/zenodo.1211533](https://doi.org/10.5281/zenodo.1211533)) [@cannoodt_goldstandardsinglecell_].

```{r}
add_stable(
  read_rds(figure_file("datasets.rds","2-dataset_characterisation/2-real"))[[params$table_format]], 
  "datasets", 
  "Real datasets used in this study."
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "samplers.rds",
    "2-dataset_characterisation/1-synthetic")
  )[[params$table_format]], 
  "samplers", 
  "Distributions from which each parameter in the thermodynamic model was sampled."
)
```

@@

```{r}
add_sfig(
  figure_file("edge_flip_comparison.svg", experiment_id="3-metric_characterisation/4-edge_flip"), 
  "edge_flip_comparison", 
 "Edge flip score values for common trajectory topologies",
  ""
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "qc_checks.rds",
    "4-method_characterisation"
  ))[[params$table_format]], 
  "qc_checks", 
  "Scoring scheme for method quality control. Each quality aspect was given a weight based on how many times it was mentioned in a set of articles discussing best practices for tool development."
)
```

@@

```{r "supplementary", echo=FALSE, results='asis'}
if (params$table_format == "html") 
  source(paste0(dynalysis::get_dynalysis_folder(), "/analysis/paper/sfigs_stables.R"))
write_rds(lst(refs, sfigs, stables), derived_file("sfigs_stables.rds", "paper"))
write_rds(figs, derived_file("figs.rds", "paper"))
```
